Sampling requirements to detect ecosystem change

### 2020-10-07 R/manuscript_draft.Rmd
work for answers to reviewer

pacman::p_load(patchwork, stickylabeller, ggplot2, dplyr, wesanderson, reshape, cowplot, vegan, ecp)
```

@@

In a previous study, Cladoceran assemblage was reconstructed continuously with a 1-cm subsampling resolution (2-3 years resolution), from a 74-cm sediment core covering the 1816(±26)—2010 time period [@Bruel2018]. Our objective was to evaluate whether the same changepoints could be identified using fewer samples.\linelabel{text:fewer} In this previous study, the variability in the community was summarized into independent axis using Detrended Component Analysis [@Hill1980]. Changepoints were then detected on the first component (46% of the total variability) in years 1926, 1946, and 1983. We defined these as the "true" changepoints given they came from an analysis of the complete data. The 1983 changepoint was the largest in magnitude, hence the changepoint we sought to find with our method. We also identified second and third changepoints (Fig. S3).

@@

\linelabel{text:detection2} \textcolor{red}{To show the generality of our approach}, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig. S4). We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig. S5). The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig. \ref{fig:case_study_number_sample}c). The implication is that a very limited number of processed sample can rapidly and efficiently be used to narrow down different questions on a same dataset. \textcolor{red}{In addition, we examined the same abundance time series, but with linear and generalized additive models (GAMs)}. In line with findings from} @White2019e, \textcolor{red}{we found that the flexibility of GAMs allow the use of less samples (Figs. S6, S7).}

@@

Our approach is applicable to a wide range of systems and questions, but it does have limitations. When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample, and analyzing all samples to obtain a whole picture may be preferred. We note that if resources need to be saved by collecting less samples in the first place, then regular sampling performs better than random sampling (Figs. \ref{fig:simulation_approaches_plot}, S1, S2). \linelabel{red}{text:limits} \textcolor{red}{Our specific iterative approach for detecting changepoints is also not appropriate for systems where changepoints are not expected. Instead, more flexible models (e.g., GAMs) might be more appropriate (Fig. S7).} Another example where our method is less useful is when addressing questions that require a continuous time series, or at least a regular sampling interval. For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016].

@@

ERW and RB conceived the ideas and designed methodology; RB collected the data; ERW designed the simulations; RB wrote the R code that simulates the three sampling approaches; ERW and RB analysed the data; Both authors contributed critically to the drafts and gave final approval for publication

@@

mypal <- wes_palette("GrandBudapest2", max(vr_out_order$count),type = "continuous")
#mypal <-  c("#4C4F54", "#D57A5D", "#AE5654", "#D18437", "#9EC6DD")

>>>>

pacman::p_load(patchwork, stickylabeller, ggplot2, dplyr, wesanderson, reshape, cowplot, vegan, ecp, ggthemes)
```

@@

In a previous study, Cladoceran assemblage was reconstructed continuously with a 1-cm subsampling resolution (2-3 years resolution), from a 74-cm sediment core covering the 1816(±26)—2010 time period [@Bruel2018]. Our objective was to evaluate whether the same changepoints could be identified using fewer samples.\linelabel{text:fewer} In this previous study, the variability in the community was summarized into independent axis using Detrended Component Analysis [@Hill1980]. Changepoints were then detected on the first component (46% of the total variability) in years 1926, 1946, and 1983. We defined these as the "true" changepoints given they came from an analysis of the complete sequence The 1983 changepoint was the largest in magnitude, hence the changepoint we sought to find with our method. We also identified second and third changepoints (Fig. S3).

@@

\linelabel{text:detection2} \textcolor{red}{To show the generality of our approach}, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig. S4). We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig. S5). The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig. \ref{fig:case_study_number_sample}c). The implication is that a very limited number of processed sample can rapidly and efficiently be used to narrow down different questions on a same dataset. \textcolor{red}{In addition, we examined the same abundance time series, but with linear and generalized additive models (GAMs)}. In line with findings from @White2019e, \textcolor{red}{we found that the flexibility of GAMs allow the use of less samples (Figs. S6, S7).}

@@

Our approach is applicable to a wide range of systems and questions, but it does have limitations. When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample, and analyzing all samples to obtain a whole picture may be preferred. We note that if resources need to be saved by collecting less samples in the first place, then regular sampling performs better than random sampling (Figs. \ref{fig:simulation_approaches_plot}, S1, S2). \linelabel{text:limits} \textcolor{red}{Our specific iterative approach for detecting changepoints is also not appropriate for systems where changepoints are not expected. Instead, more flexible models (e.g., GAMs) might be more appropriate (Fig. S7).} Another example where our method is less useful is when addressing questions that require a continuous time series, or at least a regular sampling interval. For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016].

@@

RB and ERW conceived the ideas and designed methodology; RB collected the data; ERW designed the simulations; RB wrote the R code that simulates the three sampling approaches, and applied the method to the case study data; ERW and RB analysed the data; Both authors contributed critically to the drafts and gave final approval for publication.

@@

mypal = colorblind_pal()(8)
mypal = mypal[c(7,2,3,6,8)]
mypal <- wes_palette("GrandBudapest2", max(vr_out_order$count),type = "continuous")
mypal <- wes_palette("Zissou1", max(vr_out_order$count)+5,type = "continuous")
mypal[4] <- "#90b581"
mypal <- mypal[c(1,3,4,7,9)]

### 2020-10-07 R/supp_material.Rmd
work for answers to reviewer

Throughout the main text and supplementary material, we focus on detecting changepoints. However, other types of models and criteria may be more appropriate depending on the question. This is especially true for situations where changepoints are not expected apriori. To address this, we used our random and regular sampling approaches with two additional models: linear models and generalized additive models. For simple linear regression, we measured the percent difference in the estimated slope between the subsampled data and the full time series (Fig. \ref{fig:lineartrends}). This is similar to past work [@White2019e; @White2020] on detecting population trends.

```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Percent difference for the estimated slope from linear regression for the abundance time series using (a) random and (b) regular sampling. Iterative sampling could be used for linear trends, but would require additional decisions on how to add samples. \\label{fig:lineartrends}'}

@@

Similarly, for a generalized additive model, we examined the shape of the estimated curve for subsets of data compared to the full time series (Fig. \ref{fig:gamtrends}). In line with our other results, the regular sampling performed better than random sampling upon visual inspection. Both sampling approaches also visually approach the full data set with fewer samples compared to the linear models.

```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Percent difference for the estimated slope from linear regression for the abundance time series using (a) random and (b) regular sampling. Iterative sampling could be used for linear trends, but would require additional decisions on how to add samples. \\label{fig:gamtrends}'}

@@

par(mfrow=c(2,4),mar=c(0.5,0.1,1.5,0.1),oma=c(4,4,0.5,0.5))

subset_index <- sort(c(1,sample(2:(length(vr_count$Year)-1),size = 10),length(vr_count$Year)))
vr_count_subset <- vr_count[subset_index,]
gam1_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot(gam1_rand,ylim=c(-50000,50000),xaxt='n')
mtext('a. Random (n=10)',side = 3,line = 0,adj = 0)

subset_index <- sort(c(1,sample(2:(length(vr_count$Year)-1),size = 20),length(vr_count$Year)))
vr_count_subset <- vr_count[subset_index,]
gam2_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot(gam2_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n')
mtext('b. Random (n=20)',side = 3,line = 0,adj = 0)

subset_index <- sort(c(1,sample(2:(length(vr_count$Year)-1),size = 30),length(vr_count$Year)))
vr_count_subset <- vr_count[subset_index,]
gam3_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot(gam3_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n')
mtext('c. Random (n=30)',side = 3,line = 0,adj = 0)

gam4_rand <- gam(vr_count$Biomass ~ s(vr_count$Year,k=smoothing_parameter),family=gaussian)
plot(gam4_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n')
mtext('d. Random (n=74)',side = 3,line = 0,adj = 0)

@@

subset_index <- c(seq(from = 1, to = length(vr_count$Year),length.out =10),length(vr_count$Year))
vr_count_subset <- vr_count[subset_index,]
gam1_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot(gam1_rand,ylim=c(-50000,50000))
mtext('e. Regular (n=10)',side = 3,line = 0,adj = 0)

subset_index <- c(seq(from = 1, to = length(vr_count$Year),length.out =20),length(vr_count$Year))
vr_count_subset <- vr_count[subset_index,]
gam2_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot(gam2_rand,ylim=c(-50000,50000),yaxt='n')
mtext('f. Regular (n=20)',side = 3,line = 0,adj = 0)

subset_index <- c(seq(from = 1, to = length(vr_count$Year),length.out =30),length(vr_count$Year))
vr_count_subset <- vr_count[subset_index,]
gam3_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot(gam3_rand,ylim=c(-50000,50000),yaxt='n')
mtext('g. Regular (n=30)',side = 3,line = 0,adj = 0)

gam4_rand <- gam(vr_count$Biomass ~ s(vr_count$Year,k=smoothing_parameter),family=gaussian)
plot(gam4_rand,ylim=c(-50000,50000),yaxt='n')
mtext('h. Regular (n=40)',side = 3,line = 0,adj = 0)

@@

>>>>

Throughout the main text and supplementary material, we focus on detecting changepoints. However, other types of models and criteria may be more appropriate depending on the question. This is especially true for situations where changepoints are not expected _a priori_. To address this, we used our random and regular sampling approaches with two additional models: linear models and generalized additive models. For simple linear regression, we measured the percent difference in the estimated slope between the subsampled data and the full time series (Fig. \ref{fig:lineartrends}). This is similar to past work [@White2019e; @White2020] on detecting population trends.

```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Percent difference for the estimated slope from linear regression for the abundance time series using (a-d) random and (e-h) regular sampling. Iterative sampling could be used for linear trends, but would require additional decisions on how to add samples. \\label{fig:lineartrends}'}

@@

Similarly, for a generalized additive model, we examined the shape of the estimated curve for subsets of data compared to the full time series (Fig. \ref{fig:gamtrends} \textcolor{red}{the caption is not the correct one, I edited it, but check you agree with it!}). In line with our other results, the regular sampling performed better than random sampling upon visual inspection. Both sampling approaches also visually approach the full data set with fewer samples compared to the linear models \textcolor{red}{(I do not understand this sentence)}. Iterative sampling requires only 8 samples to find the changepoint, but the a larger initial number is also necessary to capture the whole dynamic.

```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 4.4,fig.cap='The fitted smooth functions for Year for different subset of the zooplankton abundance dataset for Lake Varese (IT). Subsamples using (a-d) random and (e-h) regular sampling. (i-j) shows the influence of the initial number of sample  (n2) when using the iterative method. For low initial numbers, the dynamic of the abundance is not caught. Choosing a larger initial number (n2) allows to capture the dynamic and find the real changepoint with a limited total number of samples (n). \\label{fig:gamtrends}'}

@@

par(mfrow=c(3,4),mar=c(0.5,0.1,1.5,0.1),oma=c(4,4,0.5,0.5))

# Random GAMs

subset_index <- sort(c(1,sample(2:(length(vr_count$Year)-1),size = 10),length(vr_count$Year)))
vr_count_subset <- vr_count[subset_index,]
gam1_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot.gam(gam1_rand,ylim=c(-50000,50000),xaxt='n', shade = T)
axis(1, at=vr_count_subset$Year, NA, tcl=0.3, lwd=.5)
mtext('a. Random (n=10)',side = 3,line = 0,adj = 0)

subset_index <- sort(c(1,sample(2:(length(vr_count$Year)-1),size = 20),length(vr_count$Year)))
vr_count_subset <- vr_count[subset_index,]
gam2_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot.gam(gam2_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n', shade = T)
axis(1, at=vr_count_subset$Year, NA, tcl=0.3, lwd=.5)
mtext('b. Random (n=20)',side = 3,line = 0,adj = 0)

subset_index <- sort(c(1,sample(2:(length(vr_count$Year)-1),size = 30),length(vr_count$Year)))
vr_count_subset <- vr_count[subset_index,]
gam3_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian)
plot.gam(gam3_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n', shade = T)
axis(1, at=vr_count_subset$Year, NA, tcl=0.3, lwd=.5)
mtext('c. Random (n=30)',side = 3,line = 0,adj = 0)

gam4_rand <- gam(vr_count$Biomass ~ s(vr_count$Year,k=smoothing_parameter),family=gaussian)
plot.gam(gam4_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n', shade = T)
axis(1, at=vr_count$Year, NA, tcl=0.3, lwd=.5)
mtext('d. Random (n=74)',side = 3,line = 0,adj = 0)

@@

subset_index <- c(seq(from = 1, to = length(vr_count$Year),length.out =10),length(vr_count$Year))
vr_count_subset <- vr_count[subset_index,]
gam1_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot.gam(gam1_rand,ylim=c(-50000,50000),xaxt='n', shade = T)
axis(1, at=vr_count_subset$Year, NA, tcl=0.3, lwd=.5)
mtext('e. Regular (n=10)',side = 3,line = 0,adj = 0)

subset_index <- c(seq(from = 1, to = length(vr_count$Year),length.out =20),length(vr_count$Year))
vr_count_subset <- vr_count[subset_index,]
gam2_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot.gam(gam2_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n', shade = T)
axis(1, at=vr_count_subset$Year, NA, tcl=0.3, lwd=.5)
mtext('f. Regular (n=20)',side = 3,line = 0,adj = 0)

subset_index <- c(seq(from = 1, to = length(vr_count$Year),length.out =30),length(vr_count$Year))
vr_count_subset <- vr_count[subset_index,]
gam3_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot.gam(gam3_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n', shade = T)
axis(1, at=vr_count_subset$Year, NA, tcl=0.3, lwd=.5)
mtext('g. Regular (n=30)',side = 3,line = 0,adj = 0)

gam4_rand <- gam(vr_count$Biomass ~ s(vr_count$Year,k=smoothing_parameter),family=gaussian)
plot.gam(gam4_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n', shade = T)
axis(1, at=vr_count$Year, NA, tcl=0.3, lwd=.5)
mtext('h. Regular (n=74)',side = 3,line = 0,adj = 0)

# Iterative GAMs

subset_index <- sample_iterative(vr, xcol = 1, n = 10, input_vector = F, DCA_axis = 1)
subset_index <- subset_index$matrix$index
vr_count_subset <- vr_count[subset_index,]
gam1_it <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot.gam(gam1_it,ylim=c(-50000,50000), shade = T)
mtext('i. Iterative (n=8, n2=5)',side = 3,line = 0,adj = 0)

subset_index <- sample_iterative(vr, xcol = 1, n = 20, n2 = 10, input_vector = F, DCA_axis = 1)
subset_index <- subset_index$matrix$index
vr_count_subset <- vr_count[subset_index,]
gam2_it <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot.gam(gam2_it,ylim=c(-50000,50000),yaxt='n', shade = T)
mtext('j. Iterative (n=13, n2=10)',side = 3,line = 0,adj = 0)

subset_index <- sample_iterative(vr, xcol = 1, n = 50, n2 = 30, input_vector = F, DCA_axis = 1)
subset_index <- subset_index$matrix$index
vr_count_subset <- vr_count[subset_index,]
gam3_it <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot.gam(gam3_it,ylim=c(-50000,50000),yaxt='n', shade = T)
mtext('h. Iterative (n=32, n2=30)',side = 3,line = 0,adj = 0)

gam4_it <- gam(vr_count$Biomass ~ s(vr_count$Year,k=smoothing_parameter),family=gaussian)
plot.gam(gam4_it,ylim=c(-50000,50000), shade = T)
mtext('k. Iterative (n2=74)',side = 3,line = 0,adj = 0)

@@

# Word of caution

The limits of our three sub-sampling functions are tied to the limits of the _e.divise()_ function. In a situation where the change is a linear change instead of a sudden changepoint, the function _e.divise()_ will find a changepoint in the middle of the linear decline (Fig. \ref{fig:caution1}). In such situation, iterative sampling still performs the best.

```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 6, fig.height= 8,fig.cap='Examples of two time-series where shift lasted over (a) 5 and (b) 10 observations. Building off the example time-series in (b), we tested the three sampling method in (c-d). (c-d) are same as in Fig. 1 in the main manuscript. \\label{fig:caution1}'}

mypal <- wes_palette("Chevalier1")
mypal <- wes_palette("Zissou1", max(vr_out_order$count)+5, type = "continuous")
mypal[4] <- "#90b581"
mypal <- mypal[c(1,3,4,7,9)]

ar1 <- build_time_series()
mydata <- (ar1 - min(ar1))/(max(ar1) - min(ar1))

cex_titles <- .8

# Require an output from any of the 3 functions sampling_regular, sampling_random, sampling_iterative
pol_break <- function(cpt, col=1, y1=-.2, y2=1.2) {
  if(col==1) mycol=mypal[1]
  if(col==2) mycol=mypal[2]
  x1   <- cpt$changepoint
  x2   <- cpt$matrix$index[which(cpt$matrix$index==cpt$changepoint)-1]
  pol_x <- c(x1, x2, x2, x1)
  pol_y <- c(y1, y1, y2, y2)
  polygon(pol_x, pol_y, col= adjustcolor(mycol, alpha.f = .3), density = NULL, border = NA)
}

layout(matrix(c(1,2,0,
         3,4,5,
         0,0,6,
         0,0,7), ncol=3, byrow = T))
par(mar=c(1,2,6,2))

# plot 1: shift_duration = 5 ####
ar1 <- build_time_series(shift_duration = 5, shift_time = 62)
ar1 <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
mydata <- ar1

plot(mydata, col="black", type="l", lwd=1.5, axes=F, xlab="",ylab="")
axis(1, lwd=2, at = c(-10,120))
axis(2, lwd=2, at = c(-1,2))
cp <- sample_regular(mydata, n=length(mydata), messages = F)
pol_break(cpt = cp)
lines(mydata, col="black", lwd=1.5)
text(cp$changepoint, .8, labels = "'real' \nchange point",pos = 4, col=mypal[1])
mtext("a. Example time-series with a shift \nlasting over 5 observations", side = 3, line = 2.5, adj = 0, font=1, cex=cex_titles)

# plot2: shift_duration = 10  ####
ar1 <- build_time_series(shift_duration = 10, shift_time = 35)
ar1 <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
mydata <- ar1

plot(mydata, col="black", type="l", lwd=1.5, axes=F, xlab="",ylab="")
axis(1, lwd=2, at = c(-10,120))
axis(2, lwd=2, at = c(-1,2))
cp <- sample_regular(mydata, n=length(mydata), messages = F)
pol_break(cpt = cp)
lines(mydata, col="black", lwd=1.5)
text(cp$changepoint, .8, labels = "'real' \nchange point",pos = 4, col=mypal[1])
mtext("a. Example time-series with a shift \nlasting over 10 observations", side = 3, line = 2.5, adj = 0, font=1, cex=cex_titles)

# Plot 3 -- random sampling ####
plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="")
axis(1, lwd=2, at = c(-10,120))
axis(2, lwd=2, at = c(-1,2))
cp_random <- sample_random(mydata, n=15, messages = F)
points(cp_random$matrix$index, cp_random$matrix$x, pch=4, lwd=2)
pol_break(cpt = cp, col=1)
pol_break(cpt = cp_random, col=2)
lines(mydata, col="black", lwd=1.5)
text(100,0.95,labels = paste0("n = ",cp_random$final_n), pos = 2)
mtext("c. Random sampling", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)

# Plot 4 -- regular sampling ####
plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="")
axis(1, lwd=2, at = c(-10,120))
axis(2, lwd=2, at = c(-1,2))
cp_regular <- sample_regular(mydata, n=15, messages = F)
points(cp_regular$matrix$index, cp_regular$matrix$x, pch=4, lwd=2)
pol_break(cpt = cp, col=1)
pol_break(cpt = cp_regular, col=2)
lines(mydata, col="black", lwd=1.5)
text(100,0.95,labels = paste0("n = ",cp_regular$final_n), pos = 2)
mtext("d. Regular sampling", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)

# Plot 5 -- iterative sampling (1/3) ####
plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="")
cp_iterative <- sample_regular(mydata, n=5, messages = F)
pol_break(cpt = cp, col=1)
pol_break(cpt = cp_iterative, col=2)
lines(mydata, col="black", lwd=1.5)
axis(1, lwd=2, at = c(-10,120))
axis(2, lwd=2, at = c(-1,2))
points(cp_iterative$matrix$index, cp_iterative$matrix$x, pch=4, lwd=2)
text(100,0.95,labels = paste0("n = ",cp_iterative$final_n), pos = 2)
mtext("e. Iterative sampling (initial)", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)
xleft_temp <- cp_iterative$matrix$index[which(cp_iterative$matrix$index==cp_iterative$changepoint)-1]-4 
xright_temp <- cp_iterative$changepoint+4 
par(xpd=T)
rect(xleft = xleft_temp, xright = xright_temp, ybottom = -0.1, ytop = 1.1, lty = 2)
text(xright_temp, 1, labels = "f", pos = 4)
par(xpd=F)

# Plot 6 -- iterative sampling (2/3) ####
plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="", xlim=c(xleft_temp,xright_temp))
cp_iterative1 <- sample_regular(mydata, n=5, messages = F)
cp_iterative2 <- sample_iterative(mydata, n=6, messages = F)
cp_iterative3 <- sample_iterative(mydata, n=7, messages = F)
cp_iterative4 <- sample_iterative(mydata, n=8, messages = F)
axis(1, lwd=2, at = c(-10,120))
axis(2, lwd=2, at = c(-1,2))
mypal2 <- wes_palette("GrandBudapest2")
mypal2 <- wes_palette("Zissou1", max(vr_out_order$count)+5,type = "continuous")
mypal2[4] <- "#90b581"
mypal2 <- mypal[c(1,3,4,7,9)]

points(cp_iterative1$matrix$index, cp_iterative$matrix$x, col=grey(0.5), pch=4, lwd=2)
points(cp_iterative2$matrix$index[!cp_iterative2$matrix$index%in%cp_iterative1$matrix$index], cp_iterative2$matrix$x[!cp_iterative2$matrix$index%in%cp_iterative1$matrix$index], 
       col=mypal2[1], pch=4, lwd=2)
points(cp_iterative3$matrix$index[!cp_iterative3$matrix$index%in%cp_iterative2$matrix$index], cp_iterative3$matrix$x[!cp_iterative3$matrix$index%in%cp_iterative2$matrix$index], 
       col=mypal2[2], pch=4, lwd=2)
points(cp_iterative4$matrix$index[!cp_iterative4$matrix$index%in%cp_iterative3$matrix$index], cp_iterative4$matrix$x[!cp_iterative4$matrix$index%in%cp_iterative3$matrix$index], 
       col=mypal2[3], pch=4, lwd=2)
legend("topright", legend = c("t=0", "t=1", "t=2", "t=3"),
       col = c(grey(0.5), mypal2), pch=4, lwd=2, bty="n", lty=NA)
mtext("f. Zoom from e\n(slowly adding samples)", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)

# Plot 7 -- iterative sampling (3/3) ####
plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="")
cp_iterative <- sample_iterative(mydata, n=15, messages = F)
pol_break(cpt = cp, col=1)
pol_break(cpt = cp_iterative, col=2)
lines(mydata, col="black", lwd=1.5)
axis(1, lwd=2, at = c(-10,120))
axis(2, lwd=2, at = c(-1,2))
points(cp_iterative$matrix$index, cp_iterative$matrix$x, pch=4, lwd=2)
text(100,0.95,labels = paste0("n = ",cp_iterative$final_n), pos = 2)
mtext("g. Iterative sampling (final)", side = 3, line = 1, adj = 0, font=1, cex=cex_titles)

Another issue emerge in time series with no changepoint. The first case is when the time-series follows a declining trend. In some situations, depending on the noise and the trend, a changepoint will still be detected (Fig. \ref{fig:caution2}). Again, this would be the case even with a continuous sampling, but the iterative method can falsely reinfore this (by converging to a changepoint, and hiding the general trend). In this scenario, increasing the initial number of samples analyzed (~30% of the initial samples) would allow to roughly model the behavior of the dataset (Fig. \ref{fig:gamtrends}).

Overall, while it is important for potential users to be aware of the behavior of the code, community changes would more often lead to larger restructurations (alternative stable states, @Bruel2018). We caution in general against using too few samples for a dataset or a type of data we know too little about.

```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 6,fig.cap='Examples of two time-series where shift lasted over (a) 5 and (b) 10 observations. Building , off the example time-series in (b), we tested the three sampling method in (c-d). (c-d) are same as in Fig. 1 in the main manuscript. \\label{fig:caution2}'}

par(mfrow=c(2,2))
# Plot 1: no shift, declining trend ####
ar1 <- -0.0001*x + 1 + rnorm(length(x),0,0.001)
ar1 <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
mydata <- ar1

plot(mydata, col="black", type="l", lwd=1.5, axes=F, xlab="",ylab="")
axis(1, lwd=2, at = c(-10,120))
axis(2, lwd=2, at = c(-1,2))
cp <- sample_regular(mydata, n=length(mydata), messages = F)
if(!is.na(cp$changepoint)) pol_break(cpt = cp)
lines(mydata, col="black", lwd=1.5)
if(!is.na(cp$changepoint)) {
  text(cp$changepoint, .8, labels = "change point \ndetected",pos = 4, col=mypal[1])
} else {
    text(3,.95, labels = "No changepoint", adj = 0, col=mypal[1])
  }
mtext("a. Example time-series with no shift, only a declining trend. \nNo changepoint should have been detected.", side = 3, line = 1.5, adj = 0, font=1, cex=cex_titles)

# Plot 2: no shift, trend, subsample ####
plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="")
axis(1, lwd=2, at = c(-10,120))
axis(2, lwd=2, at = c(-1,2))
cp_regular <- sample_regular(mydata, n=15, messages = F)
points(cp_regular$matrix$index, cp_regular$matrix$x, pch=4, lwd=2)
#pol_break(cpt = cp, col=1)
pol_break(cpt = cp_regular, col=2)
text(100,0.95,labels = paste0("n = ",cp_regular$final_n), pos = 2)
mtext("b. Subsampling a time-series with no shift and a trend. \nNo changepoint should have beend detected.", side = 3, line = 1.5, adj = 0, font=1, cex=cex_titles)

# Plot 2: no shift, no trend ####
ar1 <- -0.000001*x + 1 + rnorm(length(x),0,0.001)
ar1 <- (ar1 - min(ar1))/(max(ar1) - min(ar1))
mydata <- ar1

plot(mydata, col="black", type="l", lwd=1.5, axes=F, xlab="",ylab="")
axis(1, lwd=2, at = c(-10,120))
axis(2, lwd=2, at = c(-1,2))
cp <- sample_regular(mydata, n=length(mydata), messages = F)
if(!is.na(cp$changepoint)) pol_break(cpt = cp)
lines(mydata, col="black", lwd=1.5)
if(!is.na(cp$changepoint)) {
  text(cp$changepoint, .8, labels = "'real' \nchange point",pos = 4, col=mypal[1])
} else {
    text(3,.95, labels = "No changepoint", adj = 0, col=mypal[1])
  }
mtext("c. Example time-series with no shift and no trend.\nNo changepoint should have beend detected.", side = 3, line = 1.5, adj = 0, font=1, cex=cex_titles)

plot(mydata, col=adjustcolor("grey", alpha.f = .4), type="l", lwd=1.5, axes=F, xlab="",ylab="")
axis(1, lwd=2, at = c(-10,120))
axis(2, lwd=2, at = c(-1,2))
cp_regular <- sample_regular(mydata, n=15, messages = F)
points(cp_regular$matrix$index, cp_regular$matrix$x, pch=4, lwd=2)
pol_break(cpt = cp_regular, col=2)
text(100,0.95,labels = paste0("n = ",cp_regular$final_n), pos = 2)
mtext("d. Subsampling a time-series with no shift and no trend. \nNo changepoint detected.", side = 3, line = 1.5, adj = 0, font=1, cex=cex_titles)
